---
title: "Incorporating VoC scenarios in covidseir model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting COVID-19 case data with covidseir}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.asp = 0.62,
  fig.width = 5
)
options(future.rng.onMisuse = "ignore")
```

The covidseir R package can be installed from <https://github.com/seananderson/covidseir>.

```{r setup, message=FALSE, warning=FALSE}
library(covidseir)
library(dplyr)
library(ggplot2)
ymd <- lubridate::ymd
```

## VoC Introduction

The approach to modeling VoC in [covidseir](https://github.com/seananderson/covidseir) is
to establish a period where the effective transmission rate is increasing linearly
until VoC become a the dominant strain. The effective transmissibility can is then
expressed as a piece-wise linear function with the following definition:

$$
\beta(t) = \begin{cases}
 R_0 & t < t_0,\\
 R_0 + R_0(rr - 1)\frac{t - t_0}{t_1 - t_0} & t_0 \leq t < t_1, \\
 rrR_0 & t \geq t_1.
\end{cases}
$$

Where $R_0$ is the basic reproduction ratio of the wild-type, $rrR_0$ is the basic
reproduction ratio of the variant. $t_0$ is the time of community establishment of
the variant and $t_1 - t_0$ is the time to dominance. 

To illustrate how to incorporate this into fitting consider the following faked data,

```{r cases}
values <- c(330, 745, 382, 866, 481, 668, 494, 275, 584, 514, 321, 745, 375, 301, 
           531, 590, 319, 572, 511, 699, 569, 613, 672, 397, 582, 845, 565, 497, 
           613, 984, 725, 603, 627, 732, 518, 889, 929, 821, 1086, 1152, 1211, 
           1319, 1371, 1483, 1519, 1999, 1610, 2666, 2869, 2898)
ts <- seq(1,length(cases))

dat <- dplyr::tibble(value=values,t=ts)

dat %>% ggplot(aes(x=t,y=value)) +
  geom_line()
```

Let's model a single period where the fraction distancing doesn't change and so
the change in cases in time is due to VoC.

We also create the data for the VoC here including whether they are `present` 
during the time-period, the time of `establishment` in the community, the time
to dominance of the new variant `time_to_dominance`, and the relative reproduction
ratio `rr`. In this example we fix establishment at time 0 and assume 49 days to
dominance with a 50% increase in transmission (`rr = 1.5`)

```{r setup params}
f_seg <- c(0,rep(1,length(cases) - 1))

samp_frac <- rep(0.37, length(cases))

voc_pars <- c(present = TRUE, establishment = 0,
             time_to_dominance = 49, rr = 1.5)
```


We'll fit the model with the above parameters (and further assume there is no
ramp in period where social distancing rates are changing).

As we're starting part way through the pandemic we'll also need to update the
initial number of infected cases.

```{r fit1, warning=FALSE}
fit <- covidseir::fit_seir(
  daily_cases = values,
  samp_frac_fixed = samp_frac, 
  f_seg = f_seg,
  e_prior = c(0.8, 0.05),
  use_ramp = FALSE,
  f_prior = c(0.4, 0.2),
  R0_prior = c(log(2.6), 0.2),
  i0_prior = c(log(8), 1),
  N_pop = 5.1e6, # BC population
  voc_pars = voc_pars,
  iter = 500, # number of posterior samples
  fit_type = "optimizing" # for speed only
)
```

```{r print-fit}
print(fit)
```

# Visualizing the model fit

If you would like, you can choose to use parallel processing for the projections:

```{r future, eval=FALSE}
future::plan(future::multisession)
```

Now we will take the fitted model and calculate the corresponding model predictions. This can be slow, especially with future projections, and so we only use the first 50 posterior samples for this quick vignette example:

```{r proj}
proj <- covidseir::project_seir(fit, iter = 1:2,parralel=TRUE)
proj
```

Then we will take the posterior samples, re-sample 20 times from the negative binomial observation model to generate smoother predictions, and transform the output into a tidy data frame for plotting:

```{r tidy-proj}
tidy_proj <- covidseir::tidy_seir(proj, resample_y_rep = 20)
tidy_proj
```

For plotting, we need to join a date column back on, since our projections only have a column for a numeric day. Here, we do that by making a look-up-table (`lut`).

```{r join-date1}
first_day <- min(dat$date)
last_day <- 300 # how many days to create dates for
lut <- dplyr::tibble(
  day = seq_len(last_day),
  date = seq(first_day, first_day + length(day) - 1, by = "1 day")
)
tidy_proj <- dplyr::left_join(tidy_proj, lut, by = "day")
dplyr::glimpse(tidy_proj)
```

You can plot the output however you would like; however, the covidseir package includes a built-in basic plotting function. The `value_column` and `date_column` must both be columns present in the projection and observed data.

```{r plot-projection}
covidseir::plot_projection(tidy_proj, obs_dat = dat,
  value_column = "value", date_column = "date")
```
